Social Participation Ontology: community documentation, enhancements and use examples

### 2015-09-22 plosOne/plos.tex
finished standardizing figure names and some of section 4

An inspection of \vcps\ also yielded a contradiction present in Figure~\ref{fig:diaorig}: a role cannot execute, initiate or support a cause. These are done by the social actor. Therefore, to maintain \vcps\ directions and obtain a sound ontology, {\tt ops:Role} was removed from \ops\ in preliminary versions. The result can be seen directly in Figures~\ref{fig:v1} and~\ref{fig:owlCC}.

@@

Upper ontologies usage with \ops\ is under development and should receive a dedicated article, as possibilities should be inspected carefully. Pertinent and already used as upper ontologies for \ops\ are \foaf~\cite{foaf} (for linking and describing people and things they do) and \bfo~\cite{bfo} (``designed for use in supporting information retrieval, analysis and integration in scientific and other domains'' as stated on their documentation). Properties were not related to upper ontologies as reasonable relations are still being searched for. Table~\ref{ospClasses} exposes \ops\ classes and respective upper ontologies classes.

@@

\section{\ops\ utility}\label{ospUtil}
\opsi\ is meant to be useful. First, as a systematization of what is social participation to Latin America groups, as conceived by \vcps. Second, as a mean to ease linked data, and enable integration of various instances for social participation. An indicative of this pertinence is \opa, \ocd, \ontologiaa, and \obs, ontologies that already uses \ops\ as upper ontology~\cite{pnud5}.

@@

All \ops\ classes and properties \uri s are accessible via \http. A \pubby\ instance delivers information like name, labels and relations to other classes and properties. As an example, the \uri\ \url{http://purl.org/socialparticipation/ops/SocialActor} returns information about this class and all subclasses it is related to, as shown in Figure~\ref{fig:deref}.

\begin{figure}[!h]
    \centering
%    \includegraphics[width=\columnwidth]{figs/pubbyDer}
    \caption{Dereferencing an \ops\ class: the \uri\ is also an \urll, which, reached by \http, returns information for the user as shown. Also, if the client is not a browser, but a crawler or a linked data application, \pubby\ delivers plain \rdf, not user-friendly \html.}
    \label{fig:deref}
\end{figure}

@@

\begin{figure}
    \centering
%        \includegraphics[width=0.9\columnwidth]{figs/endpoint}
    \caption{Scheme of the common use of ontologies for multiple databases integration. A user or application reaches a \sparql\ endpoint. This endpoint, through ontologies, delivers data from one or more databases. Nowadays, the most usual is to find only one database available at an endpoint, and this database is usually duplicated and not synchronized, but available as a (converted) triplestore. Even so, it is possible to access multiple ontologies and it is desirable that the databases have synchronous access, i.e. without need to convert data to triples beforehand.}
    \label{endpoint}
\end{figure}

@@

%\begin{sidewaysfigure*}
    \begin{figure}
%    \vspace{24.5em}
%    \hspace*{-3.5em}
%        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
        \caption{A taxonomic diagram of an example expanded instance of \ops. {\tt ops:PaidExecutor} is defined as being the subject of a \texttt{ops:receivesFrom} relation with a \texttt{ops:SocialActor}; \texttt{ops:DownloadedMob} is defined by being the subject of a \texttt{ops:convoquedBy} relation with a \texttt{ops:SocialNetwork}. \owl\ code is online for live editing~\cite{owlExp}. The new classes added to \ops\ are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
    \end{figure}
%\end{sidewaysfigure*}

@@

\begin{table*}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{New classes considered for expansion of the \ops. The taxonomic organization of these classes, within \ops\ can be observed in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5.7cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
ErdosRenyiNetwork & SocialNetwork & also known as Poisson network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and ErdosRenyiNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, ErdosRenyiNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from any form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & an institution that does not fit previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subject of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{adjustwidth}
\end{table*}

>>>>

An inspection of \vcps\ also yielded a contradiction present in Figure~\ref{fig:diaorig}: a role cannot execute, initiate or support a cause. These are done by the social actor. Therefore, to maintain \vcps\ directions and obtain a sound ontology, {\tt ops:Role} was removed from \ops. The result can be seen directly in Figures~\ref{fig:v1} and~\ref{fig:owlCC}.

@@

Upper ontologies usage with \ops\ is under development and should receive a dedicated article, as possibilities should be inspected carefully. Pertinent and already used as upper ontologies for \ops\ are \foaf~\cite{foaf} (for linking and describing people and things they do) and \bfo~\cite{bfo} (``designed for use in supporting information retrieval, analysis and integration in scientific and other domains'' as stated on their documentation). Properties were not related to upper ontologies as reasonable relations are still being searched for. Upper ontologies classes related to each \ops\ class are also exposed in Table~\ref{ospClasses}.

@@

\section{OPS utility}\label{ospUtil}
\opsi\ is meant to be useful. First, as a systematization of what is social participation to Latin America groups, as conceived by \vcps. Second, as a mean to ease linked data, and enable integration of various instances for social participation. An indicative of this pertinence is \opa, \ocd, \ontologiaa, and \obs, ontologies that already uses \ops\ as upper ontology~\cite{pnud5}.

@@

All \ops\ classes and properties \uri s are accessible via \http. A \pubby~\cite{pubby} instance delivers information like name, labels and relations to other classes and properties. As an example, the \uri\ \url{http://purl.org/socialparticipation/ops/SocialActor} returns information about this class, as shown in Figure~\ref{fig:deref}.

\begin{figure}[!h]
    \centering
%    \includegraphics[width=\columnwidth]{figs/pubbyDer}
    \caption{Dereferencing an \ops\ class: the \uri\ is also an \urll, which, reached by \http, returns information for the user as shown. Also, if the client is not a browser, but a crawler or a linked data application, \pubby\ delivers plain \rdf, not an user-friendly \html.}
    \label{fig:deref}
\end{figure}

@@

\begin{figure}[!h]
    \centering
%        \includegraphics[width=0.9\columnwidth]{figs/endpoint}
    \caption{Scheme of the common use of ontologies for multiple databases integration. A user or application reaches a \sparql\ endpoint. This endpoint, through ontologies, delivers data from one or more databases. Nowadays, the most usual is to find only one database available at an endpoint, and this database is usually duplicated and not synchronized, but available as a (converted) triplestore. Even so, it is possible to access multiple ontologies and it is desirable that the databases have synchronous access, i.e. without need to convert data to triples beforehand.}
    \label{endpoint}
\end{figure}

@@

%\begin{sidewaysfigure*}
%\end{sidewaysfigure*}

@@

\begin{table*}[!h]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{New classes considered for expansion of the \ops. The taxonomic organization of these classes, within \ops\ can be observed in Figure~\ref{fig:owlExp}. Further information is in Section~\ref{downwards}.}
  \begin{tabular}{|l|l|p{5cm}|p{5.7cm}|}\hline
{\bf new class} & {\bf subclass of} & {\bf description} & {\bf further notes} \\\hline\hline
SocialNetwork & Organization & a social structure made up of social actors (such as individuals or organizations) and a set of dyadic ties between these actors & --//--\\\hline
FreeScaleNetwork & SocialNetwork & a network whose connectivity follows a power law & disjoint with UniformrandomNetwork and GeographicNetwork \\ \hline
ErdosRenyiNetwork & SocialNetwork & also known as Poisson network, this network sets, with equal propability, an edge between each pair of nodes & disjoint with FreeScaleNetwork and GeographicNetwork\\\hline
GeographicNetwork & SocialNetwork & a network whose connectivity is related to the distance of nodes in a metric space & Disjoint with both FreeScaleNetwork and ErdosRenyiNetwork \\\hline
SmallWorldNetwork & SocialNetwork & a network where most nodes can be reached from other nodes with few hops or steps & not disjoint with FreeScaleNetwork, ErdosRenyiNetwork and GeographicNetwork\\\hline\hline
InformalOrganization & Organization & an organization that is not formalized & disjoint with Intitution \\ \hline
Mob & InformalOrganization & a crowd of individuals & --//-- \\\hline
GiantMob & Mob & a crowd with more than 10,000 individuals & --//--\\\hline
DownloadedMob & Mob & a Mob convoqued by a network & this is a defined class, by being a network and being the subject of the relation convoquedBy with object Network \\ \hline\hline
Institution & Organization & a mechanism of social order that governs a set of individuals & disjoint with InformalOrganization \\\hline
PublicInstitution & Institution & an institution backed through public funds and controlled by the state & disjoint with Private Institution \\ \hline
PrivateInstitution & Institution & an institution backed through private fundings and controlled by private parties & disjoint with PublicInstitution \\ \hline
AcademicInstitution & Institution & an institution dedicated to education and research, which grants academic degrees & --//-- \\ \hline
NGO & Institution & a legally consituted corporation created by natural or legal people that operate independently from any form of government & --//-- \\\hline
SpuriousInstitution & Institution & an institution that holds prominent illegitimate or corrupt characteristcs & --//-- \\\hline
ExoticInstitution & Institution & an institution that does not fit previous classes or is characterized by a very unique traces & --//-- \\\hline\hline
VoluntaryExecutor & Executor & an executor that receives no formal reward for the tasks & disjoint with PaidExecutor \\ \hline
PaidExecutor & Executor & an Executor that receives formal reward for the tasks accomplished & a defined class, being an Executor and the subject of a receivesFrom predicate with SocialActor as object\\ \hline
  \end{tabular}
  \label{ospFooClass}
\end{adjustwidth}
\end{table*}

\begin{figure}[!h]
%    \vspace{24.5em}
%    \hspace*{-3.5em}
%        \includegraphics[width=1.05\textwidth]{figs/opsExpandedTax}
        \caption{A taxonomic diagram of an example expanded instance of \ops. {\tt ops:PaidExecutor} is defined as being the subject of a \texttt{ops:receivesFrom} relation with a \texttt{ops:SocialActor}; \texttt{ops:DownloadedMob} is defined by being the subject of a \texttt{ops:convoquedBy} relation with a \texttt{ops:SocialNetwork}. \owli\ code is online for live editing~\cite{owlExp}. The new classes added to \ops\ are in Table~\ref{ospFooClass}}
    \label{fig:owlExp}
    \end{figure}
